# MSWEP Model: Download and Data Processing

The **Multi-Source Weighted-Ensemble Precipitation (MSWEP)** is a sub-daily precipitation dataset with full global coverage at 0.1° resolution, spanning the period 1979 to present. The product merges gauge, satellite, and reanalysis data.

More information: <https://www.gloh2o.org/mswep/>

Try executing the chunks by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.

### Install or activate the required libraries

```{r}
#install.packages(c("googledrive", "ncdf4", "ncdf4.helpers", "raster", "reshape2", "terra", "sf"))

library(googledrive)
library(ncdf4)
library(ncdf4.helpers)
library(raster)
library(reshape2)
library(sf)
library(dplyr)
library(lubridate)
```

## 1. Enter the time window of your interest

```{r}
#Enter the start and end year you are interested in
start_year <- NA
end_year <- NA

while (is.na(start_year) || is.na(end_year)) {

  start_year <- as.numeric(readline(prompt = "Enter the start year you are interested in: "))
  
    if (is.na(start_year) || start_year < 1980) {
    print("Error: Invalid input. Please enter a valid numeric value starting from 1980.")
    next  # Restart from the beginning
  }
  
  end_year <- as.numeric(readline(prompt = "Enter the end year you are interested in: "))
  
  if (end_year < start_year) {
    print("Error: End year cannot be smaller than the start year. Please enter valid years.")
    end_year <- as.numeric(readline(prompt = "Enter the end year you are interested in: "))
  }
}

print(paste0("Input is valid. Your request will be processed from ", start_year, " to ", end_year, ". :)"))
```

## 2. Set paths and directory

In the next step, you can choose the folder where the results will be stored, and either select a shapefile representing the region (polygon) of interest or choose a CSV file containing the coordinates of interest.

```{r}
# Destination folder for downloaded files and data storage
user_wd <- readline(prompt = "Please enter your directory path: ")
user_wd <- gsub('"', '', user_wd); user_wd <- gsub('\\\\','/',user_wd)

while (!dir.exists(user_wd)) {
  print("Invalid directory. Please enter a valid one.")
  user_wd <- readline(prompt = "Please enter your directory path: ")
  user_wd <- gsub('"', '', user_wd); user_wd <- gsub('\\\\','/',user_wd)
}
print(paste("You entered a valid directory:", user_wd))

# Create the destination folder if it doesn't exist
temp_folder <- "temp"
user_wd <- file.path(user_wd, temp_folder)
user_wd <- gsub("//", "/", user_wd)

if (!dir.exists(user_wd)) {
  dir.create(user_wd)
}

# Set the path to a. shapefile or b. CSV file with coordinates 
user_choice <- readline(prompt = "Please enter 'a' to input the location of your shapefile or 'b' to for CSV with coordinates: ")

if (tolower(user_choice) == "a") {
  # Read shapefile
  shp_path <- readline(prompt = "Please enter the path to your shapefile. Example: path/to/your/folder/polygon.shp :")
  shp_path <- gsub('"', '', shp_path); shp_path <- gsub('\\\\','/',shp_path)

  while (!file.exists(shp_path)) {
    print("Invalid file path. Please enter a valid one.")
    shp_path <- readline(prompt = "Please enter the path to your shapefile. Example: path/to/your/folder/polygon.shp :")
    shp_path <- gsub('"', '', shp_path); shp_path <- gsub('\\\\','/',shp_path)
  }

  shp <- st_read(shp_path)
  print(paste("You entered a valid path for the shapefile:", shp_path))

} else if (tolower(user_choice) == "b") {
  # Read CSV
  coord_path <- readline(prompt = "Please enter the path to your CSV with coordinates. Format: two columns latitude longitude. Example: path/to/your/folder/coordinates.csv :")
  coord_path <- gsub('"', '', coord_path); coord_path <- gsub('\\\\','/',coord_path)

  while (!file.exists(coord_path)) {
    print("Invalid file path. Please enter a valid one.")
    coord_path <- readline(prompt = "Please enter the path to your CSV with coordinates. Format: two columns latitude longitude. Example: path/to/your/folder/coordinates.csv :")
    coord_path <- gsub('"', '', coord_path); coord_path <- gsub('\\\\','/',coord_path)
  }

  coord_df <- read.csv(coord_path)
  print(paste("You entered a valid path for the CSV file:", coord_path))

} else {
  cat("Invalid choice. Please enter 'a' or 'b'.\n")
}
```

## 3. Download the dataset of interest

**Connect to the SFTP Server where the MSWEP dataset is located**

The dataset will be downloaded for the assigned variable and years and stored in the pre-determined directory on your local computer.

```{r}
# Authenticate with Google Drive
drive_auth()

# Drive folder where the MSWEP daily dataset is stored
MSWEP_all <- "1Kok05OPVESTpyyan7NafR-2WwuSJ4TO9"
MSWEP_all <- drive_ls(as_id(MSWEP_all))

# List the NRT and Past variant files
nrt_folder <- drive_ls(as_id(MSWEP_all$id[2]))
past_folder <- drive_ls(as_id(MSWEP_all$id[3]))

# List the daily NetCDF files from both the NRT and Past variant files
daily_past <- drive_ls(as_id(past_folder$id[2]))
daily_nrt <- drive_ls(as_id(nrt_folder$id[2]))
#monthly_nrt <- drive_ls(as_id(nrt_folder$id[1])) # To replicate the process with monthly or hourly
#hourly_nrt <- drive_ls(as_id(nrt_folder$id[3]))

# Merge the daily NetCDF files into a single data frame
daily_df <- merge(daily_past, daily_nrt, all = TRUE) 

# Add date details from name column into new columns
date <- daily_df$name
date <- gsub(".nc", "", date)

# Extract year and day of the year
year <- substr(date, 1, 4)
day_of_year <- as.numeric(substr(date, 5, nchar(date)))

# Create a new column with the year, month, and day
daily_df <- data.frame(daily_df = daily_df)
daily_df$date <- as.Date(paste0(year, "-01-01")) + (day_of_year - 1)
daily_df <- daily_df[order(daily_df$date), ] # Sort date
daily_df$year <- year(daily_df$date)
colnames(daily_df) <- c("name", "id", "details", "date", "year")
daily_df <- daily_df[, c(4, 5, 2, 1, 3)]

# List to store downloaded file names
downloaded_files <- c()

# Dataset download
for (year in start_year:end_year) {
  # Filter the data frame for the current year
  files_for_year <- daily_df$id[daily_df$year == year]
  
  # Loop through the file IDs for the current year
  for (file_id in files_for_year) {
    # Extract date for the current file
    date <- daily_df$date[daily_df$id == file_id]
    
    # Construct the destination path for each file
    name <- paste0("MSWEP_daily_", date, ".nc")
    name <- gsub("-", "_", name)
    
    user_path <- file.path(user_wd, name)
    
    # Check if the file has already been downloaded (in case of restarting the process)
    if (name %in% downloaded_files) {
      cat("Skipping already downloaded file:", name, "\n")
      next  # Skip to the next iteration
    }
    
    # Attempt to download the file with error handling
    tryCatch(
      {
        cat("Downloading:", name, "\n")
        drive_download(as_id(file_id), path = user_path, overwrite = TRUE)
        cat("Downloaded:", name, "\n")
        
        # Add the file name to the list of downloaded files
        downloaded_files <- c(downloaded_files, name)
      },
      error = function(e) {
        cat("Error downloading:", name, "\n")
        print(e)
      }
    )
    
    Sys.sleep(2)  # Add a delay of 2 seconds between downloads to avoid rate limiting
  }
}
drive_deauth()
```

## 4. Processing of MSWEP NetCDF dataset

After running the following chunks, the results will be stored in a folder called 'results' within your specified working directory.

### 4.1. Shapefile (polygon)

#### i. Processing of average daily values for the region (polygon) of interest

```{r}
# Extract the minimum and maximum coordinates of your shapefile
bbox <- st_bbox(shp)

lat_min <- min(bbox[2])
lat_max <- max(bbox[4])
lon_min <- min(bbox[1])
lon_max <- max(bbox[3])

# Set your region of interest's latitude and longitude range
lat_range <- c(lat_min, lat_max)
lon_range <- c(lon_min, lon_max)

# List all NetCDF files in the directory
nc_files <- list.files(user_wd, pattern = ".nc", full.names = TRUE)
variable <- "precipitation"

output_daily <- data.frame()

for (i in 1:length(nc_files)) {
  
  nc <- nc_open(nc_files[i])

  lat <- ncvar_get(nc, "lat")
  lon <- ncvar_get(nc, "lon")
  units <- nc[["var"]][["precipitation"]][["units"]]

  # Find the indices of latitudes and longitudes within your region of interest
  lat_indices <- which(lat >= lat_range[1] & lat <= lat_range[2])
  lon_indices <- which(lon >= lon_range[1] & lon <= lon_range[2])

  # Data extraction from NetCDF for the region of interest
  data_var <- ncvar_get(nc, variable, start = c(lon_indices[1], lat_indices[1], 1), 
                        count = c(length(lon_indices), length(lat_indices), -1))
  
  #Extract the respective date
  date <- sub(".*MSWEP_daily_", "", nc$filename)
  date <- sub(".nc", "", date)
  date <- as.Date(date, format = "%Y_%m_%d")
  
  # Compute the daily average over the region of interest 
  p_mean <- mean(as.matrix(data_var), na.rm = TRUE)  
  
  # Join both date and mean into a same data frame
  df_var <- data.frame(day=date, p=p_mean)

  output_daily <- rbind(output_daily, df_var) # Join current year with previous result
  
  print(paste0("Results from NetCDF file ", i, " have been saved. :)"))

  nc_close(nc)
}

# Rename the columns and days based on the start and end dates
colnames(output_daily) <- c("Day", paste0("Precipitation", " [",units,"]"))

output_daily$Day <- as.character(seq.Date(as.Date(paste(start_year, "-01-01", sep = "")), 
                                              by = "days", length.out = nrow(output_daily)))

# Export the processed results as a CSV file
dir.create(file.path(user_wd, "results"))
write.csv(output_daily, file.path(user_wd, "results", "output_daily.csv"), row.names = FALSE)

```

#### ii. Processing average monthly values for the region (polygon) of interest

```{r}
month_values <- output_daily
colnames(month_values) <- c("day", "var")

month_values$day <- as.Date(month_values$day, format = "%Y-%m-%d")
month_values$month <- format(month_values$day, "%m") # Create column with correspondent month

# Compute average monthly values
output_monthly <- month_values %>%
  group_by(year = format(day, "%Y"), month) %>%
  summarise(var = mean(var, na.rm = TRUE))

colnames(output_monthly) <- c("Year", "Month", paste0("Precipitation", " [",units,"]"))

# Export the results as a CSV file
write.csv(output_monthly, file.path(user_wd, "results", "output_monthly.csv"), row.names = FALSE)
```

### 4.2. CSV with coordinates

#### i. Processing average daily values for the list of coordinates of interest

```{r}
nc_files <- list.files(user_wd, pattern = ".nc", full.names = TRUE)
colnames(coord_df) <- c("lat", "long")
variable <- "precipitation"

point_output_daily <- data.frame()

# Iterate through each NetCDF file and each row of the coordinates data frame

for (i in 1:length(nc_files)) {

  nc <- nc_open(nc_files[i])
  
  lat_nc <- ncvar_get(nc, "lat")
  long_nc <- ncvar_get(nc, "lon")

  for (j in 1:nrow(coord_df)) {

    target_lat <- coord_df$lat[j]
    target_long <- coord_df$long[j]
    units <- nc[["var"]][[variable]][["units"]]

    # Find the nearest latitude and longitude indices to the target point
    nearest_lat_index <- which.min(abs(lat_nc - target_lat))
    nearest_long_index <- which.min(abs(long_nc - target_long))

    point_data <- ncvar_get(nc, variable, start = c(nearest_long_index, nearest_lat_index, 1),
                             count = c(1, 1, -1))
    point_data <- as.vector(point_data)

  #Extract the respective date
  date <- sub(".*MSWEP_daily_", "", nc$filename)
  date <- sub(".nc", "", date)
  date <- as.Date(date, format = "%Y_%m_%d")
  
  # Join results into a same data frame
  point_df <- data.frame(
      lat = target_lat,
      long = target_long,
      date = date,
      variable = point_data
    )
  
  point_output_daily <- rbind(point_output_daily, point_df) # Join current year with previous result

  }
  print(paste0("Results from NetCDF file ", i, " have been saved. :)"))
  
  nc_close(nc)
}

# Rename the columns and days based on the start and end dates
colnames(point_output_daily) <- c("Latitude", "Longitude", "Date", paste0("Precipitation", " [",units,"]"))

# Export the results as a CSV file
dir.create(file.path(user_wd, "results"))
write.csv(point_output_daily, file.path(user_wd, "results", "point_output_daily.csv"), row.names = FALSE)
```

#### ii. Processing average monthly values for the list of coordinates of interest

```{r}
point_month_values <- point_output_daily
colnames(point_month_values) <- c("lat", "long", "day", "var")

point_month_values$day <- as.Date(point_month_values$day, format = "%Y-%m-%d")
point_month_values$month <- format(point_month_values$day, "%m") # Create column with correspondent month

# Compute average monthly values
point_output_monthly <- point_month_values %>%
  group_by(lat, long, year = format(day, "%Y"), month) %>%
  summarise(var = mean(var, na.rm = TRUE))

colnames(point_output_monthly) <- c("Latitude", "Longitude", "Year", "Month", paste0("Precipitation", " [",units,"]"))

# Export the results as a CSV file
write.csv(point_output_monthly, file.path(user_wd, "results", "point_output_monthly.csv"), row.names = FALSE)
```
